Initial environmental impacts of Piney Point wastewater discharge into Tampa Bay, Florida

### 2021-08-23 manu-draft.Rmd
starting results

Discharges from Piney Point did not occur again until 2021.  Leakages from a tear in the plastic liner of the southern holding pond (NGS-S) were suspected when water quality samples with a similar conductivity as the wastewater were detected at onsite seepage interceptor drains. The NGS-S holding pond held 480 million gallons of wastewater, as a mixture of remnant process water from phosphate production and seawater from port dredging operations.  Water quality parameters of NGS-S measured in 2019 were well above baseline conditions typical of surface waters in Tampa Bay (Table (\@ref(tab:stktab), particularly for total phosphorus (160 mg/L) and total nitrogen (230 mg/L). Due to public safety and property concerns over catastrophic failure of the holding walls, an [emergency order](https://floridadep.gov/sites/default/files/21-0323.pdf) was issued by FDEP on March 29th for HRK to begin release of wastewater from the stack into Tampa Bay to reduce physical strain on the stacks.  Unlike past discharges from the site, HRK was authorized to release wastewater through a stormwater management system (under NPDES permit) that emptied at Port Manatee.  This was done with the assumption that backwater habitats (e.g., Bishop Harbor) may be spared the impacts of additional effluent from the site.  From March 30th to April 9th, approximately 215 million gallons of wastewater were released to lower Tampa Bay.  Over this ten day period, an estimated 205 tons of nitrogen were delivered to the bay, exceeding the annual assimilative nutrient capacity of lower Tampa Bay in a matter of days [@tbep03a10].

@@

Quantitative assessments of trends included boxplot summaries, principal components analysis (PCA), Spearman rank correlations between pairs of variables, and multiple comparisons tests to assess trends over time.  Assessments were first evaluated only on total nitrogen, chlorophyll-a, and secchi disk depth as a general assessment of potential patterns in eutrophication following wastewater release. Further analyses were conducted to compare all data types together, including the entire suite of water quality data, seagrass frequency occurrences, and macroalgae frequency occurrences to identify potential mechanisms of change using seagrasses as an endpoint for evaluating potential impacts.  Data describing *K. brevis* cell concentrations were only evaluated qualitatively because the data are from event-based sampling and generally do not represent a random sample appropriate for statistical testing. Observations for each data type were typically aggregated to the weekly or monthly scale given that sampling occurred at different days over the five month period.  Spatial comparisons were based primarily on the areas identified in Figure \@ref(fig:map)a. For PCA, all variables were standardized to zero mean and unit variance so that the central tendencies and ranges of all variables were similar.  Variables with log-normal distribution were log$_10$-transformed (i.e., nutrients, chlorophyll) prior to analysis.

@@

* Figure \@ref(fig:map) map
* Figure \@ref(fig:timeline) timeline of events
* Figure \@ref(fig:wqmap) map of water quality observations
* Figure \@ref(fig:wqtrnds) boxplots by week with +/-1sd for monthly long-term
* Figure \@ref(fig:pcacors) pca and correlations of all wq variables, by area 
* Figure \@ref(fig:trnex) seagrass and macroalgae example transect
* Figure \@ref(fig:trnfrq) seagrass, macroalgae frequency occurrence estimates
* Figure \@ref(fig:allpcacors) seagrass, macralgae, wq pca and correlation matrix
* Figure \@ref(fig:redtide) red tide historic and current, fish kill reports historic and current

>>>>

Discharges from Piney Point did not occur again until 2021.  Leakages from a tear in the plastic liner of the southern holding pond (NGS-S) were suspected when water quality samples with a similar conductivity as the wastewater were detected at onsite seepage interceptor drains. The NGS-S holding pond held 480 million gallons of wastewater, as a mixture of remnant process water from phosphate production and seawater from port dredging operations.  Water quality parameters of NGS-S measured in 2019 were well above baseline conditions typical of surface waters in Tampa Bay (Table \@ref(tab:stktab)), particularly for total phosphorus (160 mg/L) and total nitrogen (230 mg/L). Due to public safety and property concerns over catastrophic failure of the holding walls, an [emergency order](https://floridadep.gov/sites/default/files/21-0323.pdf) was issued by FDEP on March 29th for HRK to begin release of wastewater from the stack into Tampa Bay to reduce physical strain on the stacks.  Unlike past discharges from the site, HRK was authorized to release wastewater through a stormwater management system (under NPDES permit) that emptied at Port Manatee.  This was done with the assumption that backwater habitats (e.g., Bishop Harbor) may be spared the impacts of additional effluent from the site.  From March 30th to April 9th, approximately 215 million gallons of wastewater were released to lower Tampa Bay.  Over this ten day period, an estimated 205 tons of nitrogen were delivered to the bay, exceeding the annual assimilative nutrient capacity of lower Tampa Bay in a matter of days [@tbep03a10].

@@

Quantitative assessments of trends included boxplot summaries, principal components analysis (PCA), Spearman rank correlations between pairs of variables, and multiple comparison tests to assess trends between months.  Assessments were first evaluated only on total nitrogen, chlorophyll-a, and secchi disk depth as a general analysis of potential patterns in eutrophication following wastewater release. Further analyses were conducted to compare all data types together, including the entire suite of water quality data, seagrass frequency occurrences, and macroalgae frequency occurrences to identify potential mechanisms of change using seagrasses as an endpoint for evaluating potential impacts.  Data describing *K. brevis* cell concentrations were only evaluated qualitatively because the data are from event-based sampling and generally do not represent a random sample appropriate for statistical testing. Observations for each data type were typically aggregated to the weekly or monthly scale given that sampling occurred at different days over the five month period.  Spatial comparisons were based primarily on the areas identified in Figure \@ref(fig:map)a. For PCA, all variables were standardized to zero mean and unit variance so that the central tendencies and ranges of all variables were similar.  Variables with log-normal distribution were log$_10$-transformed (i.e., nutrients, chlorophyll) prior to analysis.

@@

A general narrative of events in Tampa Bay following release of wastewater from Piney Point on March 30th is shown in Figure \@ref(fig:timeline).  These events are supported by data summaries and results of statistical tests described below.  After the the discharge stopped on April 9th, an initial phytoplankton response was observed near Piney Point peaking around mid-April (Area 1, Figure \@ref(fig:wqtrnd)b).  Taxa from the Bacillariophyta phylum (diatoms) were dominant in April, with a maximum chlorophyll concentration of 265 ug/L, although median concentrations for each week in April were less than 10 ug/L. The inition diatom bloom did not perist past April.  On April 20th, *K. brevis* was first observed near Anna Maria Sound at the mouth of Tampa Bay and reached bloom concentrations by May 23rd, although observations were limited to lower Tampa Bay.  Also during May, increased observations of *Lyngbya sp.* (a macroalgal cyanobacteria)

* Figure \@ref(fig:timeline) timeline of events
* Figure \@ref(fig:wqmap) map of water quality observations
* Figure \@ref(fig:wqtrnds) boxplots by week with +/-1sd for monthly long-term
* Figure \@ref(fig:pcacors) pca and correlations of all wq variables, by area 
* Figure \@ref(fig:trnex) seagrass and macroalgae example transect
* Figure \@ref(fig:trnfrq) seagrass, macroalgae frequency occurrence estimates
* Figure \@ref(fig:allpcacors) seagrass, macralgae, wq pca and correlation matrix
* Figure \@ref(fig:redtide) red tide historic and current, fish kill reports historic and current
